\Large\textsc{Regression And Resampling Techniques \\ FYS-STK4155 \\ Assignment One}

### 2019-10-06 assignment_one/article/main.tex
Done?

\usepackage[]{mathtools}
\usepackage[]{amssymb}
%\usepackage[]{libertine}
%\usepackage[]{libertinust1math}
\usepackage[]{mathpazo}
\usepackage[]{eulervm}
\usepackage[T1]{fontenc}
\usepackage[]{microtype}
\usepackage[]{geometry}
\usepackage[]{amsthm}
\usepackage[]{thmtools}
\usepackage[usedvipsnames]{xcolor}
\usepackage[]{bm}
\usepackage[]{amsmath}
\usepackage[]{commath}
\usepackage[]{todonotes}
\usepackage[]{graphicx}
\usepackage{titling}
\usepackage{mdframed}
\usepackage[]{FiraMono}
\usepackage[]{FiraSans}
\usepackage[]{siunitx}

\usepackage[backend=biber, style=alphabetic, url=false, doi=false]{biblatex}
\addbibresource{../../FYS-STK4155.bib}
\usepackage[noabbrev, capitalize]{cleveref}

\declaretheorem[style=remark, name=Remark,
mdframed={
}]{remark}
\declaretheorem[style=remark, name=Assumption,
mdframed={
}]{assumption}

@@

\newcommand{\x}{\bm{x}}
\newcommand{\C}{\mathbb{C}}
\newcommand{\X}{\bm{X}}
\newcommand{\diag}[1]{\mathrm{diag}(#1)}
\newcommand{\MSE}[1]{\mathrm{MSE}(#1)}
\newcommand{\R}[1]{\mathrm{R}^2(#1)}
\newcommand{\expect}{\mathbb{E}}
\DeclareMathOperator{\var}{Var}
\DeclareMathOperator{\cov}{Cov}
\newcommand{\OLS}{\mathrm{OLS}}
\DeclareMathOperator{\SD}{SD}
\newcommand{\ridge}{\mathrm{ridge}}
\newcommand{\lasso}{\mathrm{lasso}}
\newcommand{\python}{\textsf{Python}{}}
\newcommand{\sgn}{\,\mathrm{sign}{}}

\newcommand{\mat}[1]{\bm{#1}}
\newcommand{\y}{\bm{y}}
\newcommand{\data}{\bm{\Omega}}
\newcommand{\cost}{\mathcal{C}}
\newcommand{\N}{\mathcal{N}}
\newcommand{\basis}{\mathcal{B}}
%\newcommand{\set}[1]{\left\{\, #1 \right\}}

@@

  \begin{abstract}
	  In this project, we implement and compare three different variants of
	  linear regression. Namely, ordinary least squares (OLS), Ridge
	  regression, and Lasso regression. We test the methods on terrain data
	  taken from a region near Stavanger, Norway. Initial implementation
	  tests are done on an artificial data-set based on noisy samples of
	  the Franke function. Statistical quantities pertaining to the various
	  model functions are computed and compared. We also give an
	  introduction to the theory and motivation behind the three methods.
  \end{abstract}

@@

  \paragraph{Performance metrics.}
  In order to evaluate the how accurately the model function \( f(\data, \beta)
  \) captures trends in the target data \( \y \), a few standard performance
  metrics are  used.  Firstly, the \emph{mean squared error}:
  \begin{equation}
    \MSE{\y, \hat{\y}} \coloneqq \frac{1}{n} \sum_{i=1}^n (y_i - \hat{y}_i)^2
  \end{equation}
  which simply averages the squared error over all samples and estimates.
  Secondly, we have the \emph{coefficient of determination} or
  \emph{\(R^2\)-score}:
  \begin{equation}
    \R{\y, \hat{y}} \coloneqq 1 - \frac{\sum_{i=1}^n (y_i -
    \hat{y}_i)^2}{\sum_{i=1}^n (y_i - \overline{y})^2}.
  \end{equation}
  The \(R^2\)-score measures much of the variation in \( \y \) which can be
  attributed to a simple linear relation between \( \x \) and \( \y \). It is a
  ratio, thus a value of one tells us that all the variation in the data can be
  attributed to an approximate linear relationship between the data \(\x\) and
  the response variable \( \y \). A value of zero indicates that a non-linear
  model may be preferrable.

@@

  \subsection{The bias-variance tradeoff} 
  \label{sub:bias_variance} 
  Since we are attempting to estimate a true distribution based on finite
  samples, there are certain statistical properties it is worthwile discussing.
  We follow the derivation from \autocite{mehtaHighbiasLowvarianceIntroduction2019}.  Recall
  \cref{as:general_linear} which states that our true data is generated from a
  noisy model
  \begin{equation}
    y = f(\x) + \varepsilon
  \end{equation}
  with \( \varepsilon \sim \N(0, \sigma^2) \).  We select a finite sample from
  the true data, which constitutes our finite data-set \( \data \).
  Furthermore, we have a model function \( f(\data, \beta) \) which depends on
  the selected data-set and the model parameters \( \beta \). Thus, our
  cost-function \( \cost \) will depend on the specific sample taken from the
  true distribution. It is of interest to examine the expected out-of-sample
  error \( \expect_{\data, \varepsilon}[\cost(\data, \beta)] \) of the model
  function over the sampled data set \( \data \) and the noise \( \varepsilon
  \). This can be decomposed into three distinct parts,
  \emph{\textcolor{Maroon}{variance}}, \emph{\textcolor{Blue}{bias}}, and
  \emph{\textcolor{Green}{noise}}:
  \begin{equation}
    \label{eq:bias_variance}
    \expect_{\data, \varepsilon}[\cost(\data, \beta)] = \frac{1}{n} \textcolor{Blue}{\sum_{i=1}^n \left(f(\x_i) - \expect_{\data}\left[\hat{y}_i\right]\right)^2}  + \frac{1}{n}\textcolor{Maroon}{\sum_{i=1}^n \expect_{\data}\big[\left(\hat{y}_i - \expect_{\data} [\hat{y}_i]\right)^2\big]} + \textcolor{Green} \sigma^2.
  \end{equation}
  See \cref{ap:derivation} for the full derivation. The bias-term measures the
  deviation of the expectation of the model function \(f(\data, \beta)\) from
  the true value \( f \), while the variance-term measures the fluctuations
  arising due to using different samples of the true data.

@@

@@

\begin{figure}
    \centering
    \subbottom{%
      \includegraphics[width=0.45\linewidth]{images/OLS_MSE_score_crossval.pdf}
    }
    \subbottom{%
      \includegraphics[width=0.45\linewidth]{images/OLS_R2_score_crossval.pdf}
    }
    \caption{Ordinary least squares regression on samples from the Franke
	    function with \( N = 100 \) points in each parameter direction,
	    yielding a total data-size of \( N^2 = 10000 \) points.  We have
	    used \( K \)-fold cross validation with \( K = 10 \).  The mean
	    squared error (left) and the \(R^2\)-score (right) as a function of
	    the polynomial degree \( d \). The values are displayed both for
	    the training data (solid) and the test data (dashed). The noise is
    distributed as \( \N(0, \tfrac{5}{100})\). Here we clearly see the effect
    of the bias-variance tradeoff. Note as the model complexity increases
    (higher polynomial degree), the model at some points overfits on the
    training data, yielding bad generalization scores, as signified by the
    sharp increase in \( \mathrm{MSE}_{\mathrm{test}} \).}
    \label{fig:ols_scores}
  \end{figure}

@@

  We sample \( N = 1000 \) points in each parameter direction and compute the
  corresponding \( N^2 \) function values of the Franke function both with and
  without added noise. The noise was selected to be distributed as \( \N(0,
  1) \) with a signal-to-noise ratio \( \alpha = 0.05\).

@@

  In addition to the evaluation metrics, we also compute confidence intervals
  for each model parameter. These are displayed in \cref{tab:ols_confidence}
  for each polynomial degree up to and including \( d = 5 \).

  \begin{table}
		\begin{tabular}{llllllr}
	\toprule
	{\(d\)} &    2 &         3 &         4 &         5 \\
	\midrule
	$1$       & \num{5.315e-05} & \num{6.349e-05} & \num{7.072e-05} & \num{6.696e-05} \\[1em]
	$x$       & \num{4.673e-04} & \num{1.682e-03} & \num{4.398e-03} & \num{8.402e-03} \\
	$y$       & \num{4.673e-04} & \num{1.683e-03} & \num{4.396e-03} & \num{8.400e-03} \\[1em]
	$x^2$     & \num{3.686e-04} & \num{6.458e-03} & \num{4.694e-02} & \num{1.959e-01} \\
	$x y$     & \num{2.948e-04} & \num{4.106e-03} & \num{2.873e-02} & \num{1.195e-01} \\
	$y^2$     & \num{3.686e-04} & \num{6.460e-03} & \num{4.692e-02} & \num{1.959e-01} \\[1em]
	$x^3$     &       {} & \num{2.575e-03} & \num{8.868e-02} & \num{9.884e-01} \\
	$x^2 y$   &       {} & \num{1.985e-03} & \num{5.199e-02} & \num{5.432e-01} \\
	$x y^2$   &       {} & \num{1.985e-03} & \num{5.199e-02} & \num{5.429e-01} \\
	$y^3$     &       {} & \num{2.575e-03} & \num{8.864e-02} & \num{9.884e-01} \\[1em]
	$x^4$     &       {} &       {} & \num{2.084e-02} & \num{1.078e+00} \\
	$x^3 y$   &       {} &       {} & \num{1.585e-02} & \num{6.287e-01} \\
	$x^2 y^2$ &       {} &       {} & \num{1.528e-02} & \num{5.443e-01} \\
	$x y^3$   &       {} &       {} & \num{1.585e-02} & \num{6.285e-01} \\
	$y^4$     &       {} &       {} & \num{2.083e-02} & \num{1.078e+00} \\[1em]
	$x^5$     &       {} &       {} &       {} & \num{1.658e-01} \\
	$x^4 y$   &       {} &       {} &       {} & \num{1.253e-01} \\
	$x^3 y^2$ &       {} &       {} &       {} & \num{1.193e-01} \\
	$x^2 y^3$ &       {} &       {} &       {} & \num{1.192e-01} \\
	$x y^4$   &       {} &       {} &       {} & \num{1.253e-01} \\
	$y^5$     &       {} &       {} &       {} & \num{1.658e-01} \\
	\bottomrule
	\end{tabular}
	\caption{The variance of the model parameters \( beta \) with respect
		to the polynomial degree \( d \). The corresponding basis
		function is displayed to the left. Note in particular how the
		variance of the model parameters seems to increase with the
		polynomial degree, which is what is to be expected according to
	the \emph{bias-variance tradeoff}.}
    \label{tab:ols_confidence}
  \end{table}

@@

  We now incorporate Ridge and Lasso regression in our numerical experiments.
  We are still working with the noisy sampled data.

@@

@@

@@

  In this project we have discussed three different methods of linear
  regression.  In all three cases, we have fitted polynomial surfaces to the
  data sets. One draw-back with using polynomial surfaces is that they are
  inherently global methods. Thus, in order to capture local trends in the
  data, a very complex model is required, which (in the polynomial case) means
  surfaces of higher polynomial degree. In recent years, several attempts at
  constructing efficient local methods have been proposed. Spline-based
  (piecewise polynomial) methods in particular have been important, and several
  approaches to locally refined spline spaces have started to gain momentum.
  We note in particular that of \emph{hierarchical B-splines} (HB),
  \emph{truncated hierarchical B-splines} (THB) and \emph{locally refined
  B-splines} (LRB). These methods allow us to place more degrees of freedom in
  regions of high data-variance, thus locally increasing the model complexity.

@@

  \appendix
  \chapter{Derivation of the bias-variance-noise relationship}
  \label{ap:derivation}

@@

  Using the same trick of adding and subtracting, we may further decompose the second term as
  \begin{align}
    \begin{split}
      \expect_{\data} \big[ (f(\x_i) - \hat{y}_i)^2\big] = \left(f(\x_i) - \expect_{\data}\big[\hat{y_i}\big]\right)^2 + \expect_{\data} \left[\left(\hat{y}_i - \expect_{\data}\left[\hat{y}_i\right]\right)^2\right]
    \end{split}
  \end{align}
  \fi
  Combining the two previous equations, we obtain

>>>>

\usepackage[]{mathtools}
\usepackage[]{amssymb}
%\usepackage[]{libertine}
%\usepackage[]{libertinust1math}
\usepackage[]{mathpazo}
\usepackage[]{eulervm}
\usepackage[T1]{fontenc}
\usepackage[]{microtype}
\usepackage[]{geometry}
\usepackage[]{amsthm}
\usepackage[]{thmtools}
\usepackage[usedvipsnames]{xcolor}
\usepackage[]{bm}
\usepackage[]{amsmath}
\usepackage[]{commath}
\usepackage[]{graphicx}
\usepackage{titling}
\usepackage{mdframed}
\usepackage[]{FiraMono}
\usepackage[]{FiraSans}
\usepackage[]{siunitx}
\usepackage[]{booktabs}

\usepackage[backend=biber, style=alphabetic, url=false, doi=false]{biblatex}
\addbibresource{../../FYS-STK4155.bib}
\usepackage[]{hyperref} 
\usepackage[noabbrev, capitalize]{cleveref}

\declaretheorem[style=remark, name=Remark,
mdframed={
}]{remark}
\declaretheorem[style=remark, name=Assumption,
]{assumption}

@@

\newcommand{\x}{\bm{x}}
\newcommand{\C}{\mathbb{C}}
\newcommand{\X}{\bm{X}}
\newcommand{\diag}[1]{\mathrm{diag}(#1)}
\newcommand{\MSE}[1]{\mathrm{MSE}(#1)}
\newcommand{\R}[1]{\mathrm{R}^2(#1)}
\newcommand{\expect}{\mathbb{E}}
\DeclareMathOperator{\var}{Var}
\DeclareMathOperator{\cov}{Cov}
\newcommand{\OLS}{\mathrm{OLS}}
\DeclareMathOperator{\SD}{SD}
\newcommand{\ridge}{\mathrm{ridge}}
\newcommand{\lasso}{\mathrm{lasso}}
\newcommand{\python}{\textsf{Python}{}}
\newcommand{\sgn}{\,\mathrm{sign}{}}
\newcommand{\train}{\mathrm{train}}
\newcommand{\test}{\mathrm{test}}

\newcommand{\mat}[1]{\bm{#1}}
\newcommand{\y}{\bm{y}}
\newcommand{\data}{\bm{\Omega}}
\newcommand{\cost}{\mathcal{C}}
\newcommand{\N}{\mathcal{N}}
\newcommand{\basis}{\mathcal{B}}

\newcommand*{\github}[1]{\def\@github{\color{gray!70}#1}}
\newcommand{\githublogo}{\raisebox{-1pt}{\includegraphics[height=9pt]{images/github}}}

%\newcommand{\set}[1]{\left\{\, #1 \right\}}

@@

  \begin{abstract}
	  In this project, we implement and compare three different variants of
	  linear regression. Namely, ordinary least squares (OLS), Ridge
	  regression, and Lasso regression. We test the methods on terrain data
	  taken from a region near Stavanger, Norway. Initial implementation
	  tests are done on an artificial data-set based on noisy samples of
	  the Franke function. Statistical quantities pertaining to the various
	  model functions are computed and compared. We also give an
	  introduction to the theory and motivation behind the three methods.
          Implementation details can be found on GitHub:

	  \centering{\href{https://github.com/qTipTip/FYS-STK4155}{\githublogo \, \color{gray!50}\textit{github.com/qTipTip}}}

  \end{abstract}

@@

  \paragraph{Performance metrics.}
  In order to evaluate the how accurately the model function \( f(\data, \beta)
  \) captures trends in the target data \( \y \), a few standard performance
  metrics are  used.  Firstly, the \emph{mean squared error}:
  \begin{equation}
    \MSE{\y, \hat{\y}} \coloneqq \frac{1}{n} \sum_{i=1}^n (y_i - \hat{y}_i)^2
  \end{equation}
  which simply averages the squared error over all samples and estimates.
  Secondly, we have the \emph{coefficient of determination} or
  \emph{\(R^2\)-score}:
  \begin{equation}
    \R{\y, \hat{y}} \coloneqq 1 - \frac{\sum_{i=1}^n (y_i -
    \hat{y}_i)^2}{\sum_{i=1}^n (y_i - \overline{y})^2}.
  \end{equation}
  The \(R^2\)-score measures much of the variation in \( \y \) which can be
  attributed to a simple linear relation between \( \x \) and \( \y \).  A
  value of one tells us that all the variation in the data can be attributed to
  an approximate linear relationship between the data \(\x\) and the response
  variable \( \y \). A negative value tells us that the model performs worse
  than a horizontal line through the data.

@@

  \subsection{The bias-variance tradeoff} 
  \label{sub:bias_variance} 
  Since we are attempting to estimate a true distribution based on finite
  samples, there are certain statistical properties it is worthwile discussing.
  We follow the derivation from \autocite{mehtaHighbiasLowvarianceIntroduction2019}.  Recall
  \cref{as:general_linear} which states that our true data is generated from a
  noisy model
  \begin{equation}
    y = f(\x) + \varepsilon
  \end{equation}
  with \( \varepsilon \sim \N(0, \sigma^2) \).  We select a finite sample from
  the true data, which constitutes our finite data-set \( \data \).
  Furthermore, we have a model function \( f(\data, \beta) \) which depends on
  the selected data-set and the model parameters \( \beta \). Thus, our
  cost-function \( \cost \) will depend on the specific sample taken from the
  true distribution. It is of interest to examine the expected out-of-sample
  error \( \expect_{\data, \varepsilon}[\cost(\data, \beta)] \) of the model
  function over the sampled data set \( \data \) and the noise \( \varepsilon
  \). This can be decomposed into three distinct parts,
  \emph{\textcolor{Maroon}{variance}}, \emph{\textcolor{Blue}{bias}}, and
  \emph{\textcolor{Green}{noise}}:
  \begin{equation}
    \label{eq:bias_variance}
    \expect_{\data, \varepsilon}[\cost(\data, \beta)] = \frac{1}{n} \textcolor{Blue}{\sum_{i=1}^n \left(f(\x_i) - \expect_{\data}\left[\hat{y}_i\right]\right)^2}  + \frac{1}{n}\textcolor{Maroon}{\sum_{i=1}^n \expect_{\data}\big[\left(\hat{y}_i - \expect_{\data} [\hat{y}_i]\right)^2\big]} + \textcolor{Green} \sigma^2.
  \end{equation}
  The bias-term measures the
  deviation of the expectation of the model function \(f(\data, \beta)\) from
  the true value \( f \), while the variance-term measures the fluctuations
  arising due to using different samples of the true data.

@@

  Given a data-set \( \data = (\x_i, y_i)_{i=1}^n \), we divide it into two disjoint subset \( \data_\train \) and \( \data_\test \)
  with the purpose of optimizing our model by training on \( \data_\train \) and evaluating model performance on the ``unseen data'' of \( \data_\test \).

  Cross-validation entails repeated such splitting of the data-set into test
  and training sets, evaluating model performance on each such split, and then
  finally agregating the results to get a good estimate of model performance
  over the whole data-set. In \emph{k-fold cross-validation}, the data is
  specifically subdivided into \( k \) equal parts, where each sub-set takes on
  the role of \( \data_\test \) exactly once.

  In our implementation we have made use of \textsc{sklearn}'s \textsc{KFold} method, with \( k = 10 \).

@@

\begin{figure}
    \centering
    \subbottom{%
      \includegraphics[width=0.45\linewidth]{images/OLS_MSE_score_crossval.pdf}
    }
    \subbottom{%
      \includegraphics[width=0.45\linewidth]{images/OLS_R2_score_crossval.pdf}
    }
    \caption{Ordinary least squares regression on samples from the Franke
	    function with \( N = 100 \) points in each parameter direction,
	    yielding a total data-size of \( N^2 = 10000 \) points.  We have
	    used \( K \)-fold cross validation with \( K = 10 \).  The mean
	    squared error (left) and the \(R^2\)-score (right) as a function of
	    the polynomial degree \( d \). The values are displayed both for
	    the training data (solid) and the test data (dashed). The noise is
    distributed as \( \N(0, \tfrac{5}{100})\). Here we clearly see the effect
    of the bias-variance tradeoff. Note as the model complexity increases
    (higher polynomial degree), the model at some points overfits on the
    training data, yielding bad generalization scores, as signified by the
    sharp increase in \( \mathrm{MSE}_{\mathrm{test}} \).}
    \label{fig:ols_scores_crossval}
  \end{figure}

@@

  We randomly sampled \( N = 1000 \) points in the unit-square and compute the
  corresponding \( N \) function values of the Franke function both with and
  without added noise. The noise was selected to be distributed as \( \N{0,
  1} \) with a signal-to-noise ratio \( \alpha = 0.05 \).

@@

  In addition to the evaluation metrics, we also compute the variance for each
  model parameter.  These are displayed in \cref{tab:ols_confidence} for each
  polynomial degree from \(d = 2 \) up to and including \( d = 5 \).

  \begin{table}
		\begin{tabular}{llllllr}
	\toprule
	{\(d\)} &    2 &         3 &         4 &         5 \\
	\midrule
	$1$       & \num{5.315e-05} & \num{6.349e-05} & \num{7.072e-05} & \num{6.696e-05} \\[1em]
	$x$       & \num{4.673e-04} & \num{1.682e-03} & \num{4.398e-03} & \num{8.402e-03} \\
	$y$       & \num{4.673e-04} & \num{1.683e-03} & \num{4.396e-03} & \num{8.400e-03} \\[1em]
	$x^2$     & \num{3.686e-04} & \num{6.458e-03} & \num{4.694e-02} & \num{1.959e-01} \\
	$x y$     & \num{2.948e-04} & \num{4.106e-03} & \num{2.873e-02} & \num{1.195e-01} \\
	$y^2$     & \num{3.686e-04} & \num{6.460e-03} & \num{4.692e-02} & \num{1.959e-01} \\[1em]
	$x^3$     &       {} & \num{2.575e-03} & \num{8.868e-02} & \num{9.884e-01} \\
	$x^2 y$   &       {} & \num{1.985e-03} & \num{5.199e-02} & \num{5.432e-01} \\
	$x y^2$   &       {} & \num{1.985e-03} & \num{5.199e-02} & \num{5.429e-01} \\
	$y^3$     &       {} & \num{2.575e-03} & \num{8.864e-02} & \num{9.884e-01} \\[1em]
	$x^4$     &       {} &       {} & \num{2.084e-02} & \num{1.078e+00} \\
	$x^3 y$   &       {} &       {} & \num{1.585e-02} & \num{6.287e-01} \\
	$x^2 y^2$ &       {} &       {} & \num{1.528e-02} & \num{5.443e-01} \\
	$x y^3$   &       {} &       {} & \num{1.585e-02} & \num{6.285e-01} \\
	$y^4$     &       {} &       {} & \num{2.083e-02} & \num{1.078e+00} \\[1em]
	$x^5$     &       {} &       {} &       {} & \num{1.658e-01} \\
	$x^4 y$   &       {} &       {} &       {} & \num{1.253e-01} \\
	$x^3 y^2$ &       {} &       {} &       {} & \num{1.193e-01} \\
	$x^2 y^3$ &       {} &       {} &       {} & \num{1.192e-01} \\
	$x y^4$   &       {} &       {} &       {} & \num{1.253e-01} \\
	$y^5$     &       {} &       {} &       {} & \num{1.658e-01} \\
	\bottomrule
	\end{tabular}
	\caption{The variance of the model parameters \( \beta \) with respect
		to the polynomial degree \( d \). The corresponding basis
		function is displayed to the left. Note in particular how the
		variance of the model parameters seems to increase with the
		polynomial degree, which is what is to be expected according to the
		\emph{bias-variance tradeoff}.}
    \label{tab:ols_confidence}
  \end{table}

@@

  We now incorporate Ridge and Lasso regression in our numerical experiments.
  Both of these regression techniques rely on a hyper-parameter \( \lambda \)
  which has to be tuned to the data at hand. By plotting the estimated bias and
  variance of the  model as a function of the hyper-parameter \( \lambda \) we
  find the optimal parameter. The fits were performed on the same data as in
  the aforementioned OLS-test, and the results are presented in
  \cref{fig:bias_variance_lambda}. However, in this case we only use bi-quintic
  polynomial surfaces.

  \begin{figure}[htpb]
  	\centering
	\subbottom[Ridge regression]{%
		\includegraphics[width=0.45\linewidth]{images/bias_variance_ridge.pdf}
	}
	\subbottom[Lasso regression]{%
		\includegraphics[width=0.45\linewidth]{images/bias_variance_lasso.pdf}
	}
	\caption{Finding the optimal hyper-parameter \( \lambda \) for both
		Ridge and Lasso regression. The fits were perfomed for
		bi-quintic polynomial surfaces with noisy data sampled from the
		Franke function. As we see, the variance is fairly constant,
	until it suddenly increases. The bias on the other hand converges to
	zero. The optimal lambda is indicated by a vertical line. Note how the error is
	constant after a certain value of \( \lambda \). By looking at
	\cref{fig:ridge_lasso_beta_variance}, this is the point at which all
	Lasso-parameters are set to zero.}%
  	\label{fig:bias_variance_lambda}
  \end{figure}

  We also plot the model parameters \( \beta \) and their standard deviations
  as a function of \( \lambda \).  It is expected that the parameters \(
  \beta_\ridge\) tends to zero as \( \lambda \) increases, while the parameters
  \( \beta_\lasso\) may in addition be set equal to zero. This is verified by
  the plots in \cref{fig:ridge_lasso_beta_variance}.

  \begin{figure}[htpb]
  	\centering
	\subbottom[Ridge regression]{%
		\includegraphics[width=0.45\linewidth]{images/variance_ridge.pdf}
	}
	\subbottom[Lasso regression]{%
		\includegraphics[width=0.45\linewidth]{images/variance_lasso.pdf}
	}
	\caption{The model parameters \( \beta \) with associated standard
	deviations. We do not label each line to not clutter the plot, as its
	main purpose is to illustrate how Ridge and Lasso regression, respectively,
	penalizes model parameters. Note that for the Lasso regression, the model is
	quite sparse, and in particular, note how the \( \lambda \) where all
	parameters are zero correspond to the constant behaviour in
	\cref{fig:bias_variance_lambda}.}% 
	\label{fig:ridge_lasso_beta_variance}
  \end{figure}

  In \cref{tbl:ridge_lasso_mse_r2} we present the mean squared error and
  \(R^2\)-score for fits performed with optimal hyper parameter \( \lambda \)
  as indicated in \cref{fig:bias_variance_lambda} and compare them to an
  OLS-fit.

  \begin{table}[htpb]
  	\centering
	\caption{The MSE and \( R^2 \) score for the Franke function using
		bi-quintic polynomial surfaces and the optimal hyper parameters
		for Ridge and Lasso respectively. According to the reported
		metrics, the OLS-method performs significantly better than both
		Ridge and Lasso. The \(R^2\)-scores reported for Ridge and
	Lasso seem unreasonably low compared to OLS, which may indicate an
error in the implementation, or the authors lack of experience in the field.}
  	\label{tbl:ridge_lasso_mse_r2}
  	\begin{tabular}{llll}
		\toprule
		{} & OLS & Ridge & Lasso \\
		\midrule
		\( \mathrm{MSE} \) & \num{0.011} & \num{0.080} & \num{0.092} \\
		\( \mathrm{R}^2 \) & \num{0.879} & \num{0.140} & \num{0.005} \\
		\bottomrule
  	\end{tabular}
  \end{table}

@@

  When testing on real data, in order to keep the data-size manageable, we
  select a small patch from the large dataset.  This patch is subsequently
  normalized in order to improve numerical properties of the methods.
  Unfortunately, due to time-constraints, and numerical bugs - we only present
  the MSE and \(R^2\) scores for the various methods with fixed
  hyper-parameters lambda for Ridge and Lasso regression originally selected
  for the previous dataset.  Ideally, hyper-parameter optimzation as done in
  \cref{fig:bias_variance_lambda} should be performed for this new data-set as
  well, and further statistical quantities reported.

  The normalized patch is displayed in \cref{fig:sub_terrain}, and the performance metrics in \cref{fig:terrain_metrics}.
  \begin{figure}[htpb]
  	\centering
  	\includegraphics[width=0.8\linewidth]{images/reduced_terrain_data.pdf}
	\caption{A small patch from the larger data-set displayed in \cref{fig:terrain}. The data has been normalized.}%
	\label{fig:sub_terrain}
  \end{figure}

  \begin{figure}[htpb]
  	\centering
	\subbottom{
		\includegraphics[width=0.45\linewidth]{images/terrain_data_mse.pdf}	
	}
	\subbottom{
		\includegraphics[width=0.45\linewidth]{images/terrain_data_r2.pdf}	
	}
	\caption{The performance metrics reported for polynomial fits to the
	small terrain patch. As we can see, the Ordinary Least Squares
regression seems to be the better chocice for terrain data with high local
variance. It seems like both Ridge and Lasso struggles with finer features
going missing, as they penalize large but ``irrelevant'' model parameters.}%
	\label{fig:terrain_metrics}
  \end{figure}

@@

  In this project we have studied three kinds of linear regression, namely
  Ordinary Least Squares, Ridge, and Lasso regression.  We tested the methods
  on artificial data sampled from the Franke function, and later tested the
  methods on terrain data.

  Due to some unfortunate implementation errors and time constraints, the study
  on the terrain data were rather uncomplete, however the one result we did
  manage to get indicates that for data such as the terrain data presented
  here, the ordinary least squares method may be slightly better. All numerical
  methods were tested against \textsc{sklearn}'s methods, and the problems
  persisted, indicating rather that there is a more fundamental
  misunderstanding from the authors part.

@@

  In this project we have discussed three different methods of linear
  regression.  In all three cases, we have fitted polynomial surfaces to the
  data sets. One draw-back with using polynomial surfaces is that they are
  inherently global methods. Thus, in order to capture local trends in the
  data, a very complex model is required, which (in the polynomial case) means
  surfaces of higher polynomial degree. In recent years, several attempts at
  constructing efficient local methods have been proposed. Spline-based
  (piecewise polynomial) methods in particular have been important, and several
  approaches to locally refined spline spaces have started to gain momentum.
  We note in particujlar that of \emph{hierarchical B-splines} (HB),
  \emph{truncated hierarchical B-splines} (THB) and \emph{locally refined
  B-splines} (LRB). These methods allow us to place more degrees of freedom in
  regions of high data-variance, thus locally increasing the model complexity.

@@

  \if 0
  \appendix
  \chapter{Derivation of the bias-variance-noise relationship}
  \label{ap:derivation}

@@

  Using the same trick of adding and subtracting, we may further decompose the second term as
  \begin{align}
    \begin{split}
      \expect_{\data} \big[ (f(\x_i) - \hat{y}_i)^2\big] = \left(f(\x_i) - \expect_{\data}\big[\hat{y_i}\big]\right)^2 + \expect_{\data} \left[\left(\hat{y}_i - \expect_{\data}\left[\hat{y}_i\right]\right)^2\right]
    \end{split}
  \end{align}
  \fi
  Combining the two previous equations, we obtain
  \fi
